Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-05-30 paper/paper.tex
two BL comments

Projects like \project{The Cannon} (\citealt{ness15}; \citealt{casey17}; \citealt{ho17})
and \project{Avast} (Bedell et al.\ in preparation),
explore the extent to which our predictive models of
stars could be purely data-driven or statistical.
That is, under what circumstances could the data themselves deliver
more precise or more accurate information than any theoretical or
physics-based model?
The answer to this question is extremely context-dependent: It depends
on what data are available, how much data are available, and what
questions are being asked.
In general, data-driven models contain fewer assumptions than
physics-driven models, but they are also usually less interpretable.
However, the \gaia\ data set is ideal for thinking about purely statistical
models of stars:
The distance to a star is a latent variable, and the nearby
stars constitute a high signal-to-noise ``training set''.
These nearby stars can be used to anchor empirical models that can
then be applied to the more distant stars\BLedit{, thanks to the simple physical relations connecting the measurements (apparent magnitudes, parallaxes) to some of the physical parameters of interest (absolute magnitudes, distances).}

Here we deliver a demonstration-of-concept that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise parallaxes, without any input of the physics or numerical models of stellar
evolution, interiors, or photospheres.
That is, this is a data-driven or purely empirical \emph{photometric
  parallax} method.
A photometric parallax is a parallax (or distance) estimate based on
the photometry (colors and apparent magnitude or magnitudes) of a
star.
Fundamentally, photometric parallaxes capitalize on the fact that the
apparent magnitude of a star is a strong function of distance, and
that the absolute magnitude is a strong function of color (or surface
temperature) and evolutionary stage.

@@

Setting up the problem in more detail, we have the true samples from the underlying distribution and the noisy samples

\begin{eqnarray}
y_{\true, n} \, &=& \, m_{\true}\,x_n + b_{\true} + \Delta_n \\
y_n \, &=& \, y_{\true,n} + \epsilon_n
\label{eq:ytrue}
\end{eqnarray}

where $\Delta_n$ is drawn from the Gaussian function $\mathcal{N}(\Delta_n \given 0, t^2)$, and $\epsilon_n$ is drawn from the Gaussian function $\mathcal{N}(\epsilon_n \given 0, \sigma_n^2)$, so $\sigma_n$ is the measurement uncertainty.
These represent true and noisy draws from a distribution that is a straight line with some thickness $t$, and are visualized in the upper panels of Figure \ref{fig:toy}
The upper left shows 1024 realizations of the true values, and the upper right, as well as lower left, show 1024 realizations of the noisy values $y_n$ and their associated measurement uncertainties $\sigma_n$ as the grey points. Here we highlight 5 random points in black to highlight the method, but the same method is being applied to all the data.

To infer a pdf of the true y value for each measured y value we need a prior. For our prior we will use an estimate of the underlying true distribution that the noisy data $y_n$ are drawn from, which we build from the noisy y data themselves.
We learn the noise-deconvolved function using marginalized maximum likelihood estimation (MML), which marginalizes over the true data values (we never need to know them) and maximizes the likelihood of the data. Here our likelihood for each individual datum is

\BLedit{
\begin{eqnarray}
p(y_n \given m, b, t^2, \sigma_n^2) \ &=& \ \int \mathcal{N}(y_n \given y_{\true,n}, \sigma_n^2) \, \mathcal{N}(y_{\true,n} \given m\,x_n + b, t^2 ) \, \mathrm{d}y_{\true, n} \\
&=&  \ \mathcal{N}(y_n \given m\,x_n + b, t^2 + \sigma_n^2)
\label{eq:toyLike}
\end{eqnarray}

where we have marginalized over the true sample $y_{\true, n}$.} The resulting likelihood is a Gaussian because the noise model is a Gaussian. \BLcomment{should all these equations have $x_n$ on the right hand side? I don't have a strong opinion about that.} The likelihood of the full data set is then the product of the individual likelihoods

\begin{equation}
\prod_n\, p(y_n \given m, b, t^2, \sigma_n^2).
\label{eq:toyLikeFull}
\end{equation}

Maximizing this joint likelihood of all the data is the same as minimizing

\begin{equation}
\chi^2 = \sum_n \frac{[y_n - [m\,x_n + b]]^2}{[\sigma_n^2 + t_n^2]} + \sum_n\,\ln(2\pi\,[\sigma_n^2 + t^2])
\label{eq:chisq}
\end{equation}

>>>>

Projects like \project{The Cannon} (\citealt{ness15}; \citealt{casey17}; \citealt{ho17})
and \project{Avast} (Bedell et al.\ in preparation),
explore the extent to which our predictive models of
stars could be purely data-driven or statistical.
That is, under what circumstances could the data themselves deliver
more precise or more accurate information than any theoretical or
physics-based model?
The answer to this question is extremely context-dependent: It depends
on what data are available, how much data are available, and what
questions are being asked.
In general, data-driven models contain fewer assumptions than
physics-driven models, but they are also usually less interpretable.
However, the \gaia\ data set is ideal for thinking about purely statistical
models of stars:
The different stars observed by \gaia\ are measured at very different
signal-to-noise ratios.
A data-driven model of the stellar color--magnitude distribution can capitalize
on the precise information coming from the high signal-to-noise stars
and deliver valuable information about the low signal-to-noise stars,
which in turn will deliver precise parallax (or distance or absolute magnitude) inferences.

Here we deliver a demonstration-of-concept that shows that a
data-driven model of the \gaia\ data could in principle deliver very
precise parallax measurements, without any input of the physics or numerical models of stellar
evolution, interiors, or photospheres.
That is, this is a data-driven or purely empirical \emph{photometric
  parallax} method.
A photometric parallax is a parallax (or distance) estimate based on
the photometry (colors and apparent magnitude or magnitudes) of a
star.
Fundamentally, photometric parallaxes capitalize on the fact that the
apparent magnitude of a star is a strong function of distance, and
that the absolute magnitude is a strong function of color (or surface
temperature) and evolutionary stage.

@@

Setting up the problem in more detail, we have the true samples from the underlying distribution and the noisy samples
\begin{eqnarray}
y_{\true, n} \, &=& \, m_{\true}\,x_n + b_{\true} + \Delta_n \\
y_n \, &=& \, y_{\true,n} + \epsilon_n \quad,
\label{eq:ytrue}
\end{eqnarray}
where $\Delta_n$ is drawn from the Gaussian function $\mathcal{N}(\Delta_n \given 0, t^2)$, and $\epsilon_n$ is drawn from the Gaussian function $\mathcal{N}(\epsilon_n \given 0, \sigma_n^2)$, so $\sigma_n$ is the measurement uncertainty.
These represent true and noisy draws from a distribution that is a straight line with some thickness $t$, and are visualized in the upper panels of Figure \ref{fig:toy}
The upper left shows 1024 realizations of the true values, and the upper right, as well as lower left, show 1024 realizations of the noisy values $y_n$ and their associated measurement uncertainties $\sigma_n$ as the grey points. Here we highlight 5 random points in black to highlight the method, but the same method is being applied to all the data.

To infer a pdf of the true y value for each measured y value we need a prior. For our prior we will use an estimate of the underlying true distribution that the noisy data $y_n$ are drawn from, which we build from the noisy y data themselves.
We learn the noise-deconvolved function using marginalized maximum likelihood estimation (MML), which marginalizes over the true data values (we never need to know them) and maximizes the likelihood of the data. Here our likelihood for each individual datum is
\begin{eqnarray}
p(y_n \given m, b, t^2, \sigma_n^2) \ &=& \ \int \mathcal{N}(y_n \given y_{\true,n}, \sigma_n^2) \, \mathcal{N}(y_{\true,n} \given m\,x_n + b, t^2 ) \, \mathrm{d}y_{\true, n} \\
&=&  \ \mathcal{N}(y_n \given m\,x_n + b, t^2 + \sigma_n^2) \quad,
\label{eq:toyLike}
\end{eqnarray}
where we have marginalized over the true sample $y_{\true, n}$. The resulting likelihood is a Gaussian because the noise model is a Gaussian. \BLcomment{should all these equations have $x_n$ on the right hand side? I don't have a strong opinion about that.} The likelihood of the full data set is then the product of the individual likelihoods
\begin{equation}
\prod_n\, p(y_n \given m, b, t^2, \sigma_n^2) \quad .
\label{eq:toyLikeFull}
\end{equation}
Maximizing this joint likelihood of all the data is the same as minimizing
\begin{equation}
\chi^2 = \sum_n \frac{[y_n - [m\,x_n + b]]^2}{[\sigma_n^2 + t_n^2]} + \sum_n\,\ln(2\pi\,[\sigma_n^2 + t^2])
\quad .
\label{eq:chisq}
\end{equation}
